---
title: "How Humans See Data - Notes"
author: "John Rauser"
date: "10/10/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(scales)
library(broom)
library(viridis)

set.seed(7777)
```


# About this document

This document contains all the code I used to generate the images for my presentation How Humans See Data, first presented at [Velocity Amsterdam 2016](http://conferences.oreilly.com/velocity/devops-web-performance-eu/public/schedule/detail/54354).  This document is not meant to stand alone -- it has very little explanatory text.  It is meant to be a reference for people who want to see how I made the plots from the presentation. 

The source code is on github at [https://github.com/jrauser/writing/tree/master/how_humans_see_data](https://github.com/jrauser/writing/tree/master/how_humans_see_data).  The rendered document is available on rpubs at [http://rpubs.com/jrauser/hhsd_notes](http://rpubs.com/jrauser/hhsd_notes).

# Setup

First I set up a theme for presentations.  It uses white on black (good for large, dimly lit rooms), and lines and text that are too small/fine when rendered in a document, but which work well when rendered at 300dpi for the presnetation.  Note that I would usually not use white on black for documents, but I use this as the default theme throughout because there isn't an easy way (that I know of) to set the default color of a geom to something other than black.

```{r}
theme_base_size<-11
theme_preso<-theme_dark(base_size=theme_base_size) %+replace%
  theme(line = element_line(colour = "white", size = 0.5, linetype = 1,
                            lineend = "butt"),
        rect = element_rect(fill = "black", colour = "white", size = 0.5, linetype = 1),
        text = element_text(family = "Helvetica", face = "plain",
                            colour = "white", size = theme_base_size,
                            hjust = 0.5, vjust = 0.5, angle = 0, lineheight = 0.9, 
                            margin = unit(0.5, "lines"), debug=F),

        axis.text = element_text(color="lightgrey", size=rel(0.7)),
        axis.ticks = element_line(colour = "white", size = 0.5),

        legend.background =  element_rect(colour = NA),
        #legend.box =         NULL,
        legend.key =         element_rect(fill = "black", colour = NA),

        panel.border =       element_rect(fill = NA, colour = "darkgrey"),
        panel.border =       element_blank(),
        panel.grid.major =   element_blank(),
        panel.grid.minor =   element_blank(),
        panel.background =   element_rect(fill = "black", colour = "black"),

        strip.text.x = element_text(size=rel(0.8), color="white",
                                    margin = unit(rep(0.1, 4), "lines")),
        
        plot.background = element_rect(colour = "black", fill = "black")
)
theme_set(theme_preso)

preso_save<-function(filename, plot=last_plot(), theme=theme_preso, width=1280/300, height=720/300, dpi=300) {
  ggsave(filename, plot+theme, width=width, height=height, dpi=dpi)
}
```

# Part 1: Why visualize at all?

## A circle

Showing the data in a scatterplot makes the relationship obivous pre-attentively.

```{r}
circle<-read.delim(textConnection(
"x	y
1.972	1.236
1.112	1.994
0	1.009
0.665	1.942
0.235	0.356
0.247	1.658
1.275	1.961
0.702	0.045
1.76	0.35
1.691	0.277
1.628	1.778
1.957	1.29
0.111	0.542
0.902	0.005
0.598	0.085
1.613	1.79
1.298	1.955
0.651	1.937
1.949	1.316
0.099	0.567
0.862	0.01
0.027	0.768
0.706	1.956
1.042	1.999"))

ggplot(circle, aes(x,y))+geom_point(color="white", size=2)+coord_equal()
preso_save("~/hhsd_images/circle.png")
```

But showing it as a time-series obscures the relationship.  This is arguably worse than a simple table.

```{r}
circle_w <- circle %>% 
  mutate(n=1:nrow(circle)) %>%
  gather(var, value, x, y)
ggplot(circle_w, aes(n, value, color=var))+geom_line()+xlab(NULL)
preso_save("~/hhsd_images/circle_lines.png")
```

## Anscombe's Quartet

```{r}
quartet<-read.delim(textConnection(
"group	x	y
A	10	8
A	8	7
A	13	7.6
A	9	8.8
A	11	8.3
A	14	10
A	6	7.2
A	4	4.3
A	12	10.8
A	7	4.8
A	5	5.7
B	10	9.1
B	8	8.1
B	13	8.7
B	9	8.8
B	11	9.3
B	14	8.1
B	6	6.1
B	4	3.1
B	12	9.1
B	7	7.3
B	5	4.7
C	10	7.5
C	8	6.8
C	13	12.7
C	9	7.1
C	11	7.8
C	14	8.8
C	6	6.1
C	4	5.4
C	12	8.2
C	7	6.4
C	5	5.7
D	8	6.6
D	8	5.8
D	8	7.7
D	8	8.8
D	8	8.5
D	8	7
D	8	5.3
D	19	12.5
D	8	5.6
D	8	7.9
D	8	6.9"))
```

All four are identical on common statistical measures.

```{r}
quartet %>%
  group_by(group) %>%
  summarize(mean_x=mean(x), mean_y=mean(y))

quartet %>%
  group_by(group) %>%
  do(tidy(lm(y~x, data=.))) %>%
  select(group, term, estimate) %>%
  spread(term, estimate)

quartet %>%
  group_by(group) %>%
  do(glance(lm(y~x, data=.))) %>%
  select(group, r.squared)
```

But obivously different visually.

```{r}
ggplot(quartet, aes(x,y))+geom_point(color="white")+facet_wrap(~group)
preso_save("~/hhsd_images/anscombe.png")
```

# Part 2: Estimation

## The hierarchy

### The first rule of color is do not talk about color

Color has three separate channels.

```{r}
values<-data.frame(val=1:5, y=1)

theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank(),
                      legend.position="none")

ggplot(values, aes(val,y,fill=val)) + geom_tile() + scale_fill_gradient(low="black", high="white", limits=c(0,6)) + theme_extras
preso_save("~/hhsd_images/luminance.png", height=720/300/3, theme=theme_preso + theme_extras)
ggplot(values, aes(val,y,fill=val)) + geom_tile() + scale_fill_gradient(low="white", high="red", limits=c(0,6)) + theme_extras
preso_save("~/hhsd_images/saturation.png", height=720/300/3, theme=theme_preso + theme_extras)
ggplot(values, aes(val,y,fill=factor(val))) + geom_tile() + scale_fill_discrete() + theme_extras
preso_save("~/hhsd_images/hue.png", height=720/300/3, theme=theme_preso + theme_extras)
```


Lots of plots with the mtcars dataset.  I had to drop the font size way down to get the plots the render well for the presentation.  They are much too small for the document.  I offer my apologies.

```{r}
# Make model an actual column in the dataset.
mtcars <- mtcars %>%
  mutate(model=rownames(mtcars))
```

## Hue

The viridis hue scale.  I didn't end up using this in the presentation.  I probably should have, as it's far superior to the crappy one I made up.

```{r}
ggplot(mtcars, aes("a", model, fill=mpg))+geom_tile()+scale_fill_viridis()
ggplot(mtcars, aes("a", reorder(model, mpg), fill=mpg))+geom_tile()+scale_fill_viridis()
```

Ordered and unordered plots with hue encoding efficiency.

```{r}
theme_extras <- theme(axis.title.x=element_blank(),
                      axis.text.x=element_blank(),
                      axis.ticks.x=element_blank(),
                      axis.text.y=element_text(size=rel(0.6)))

ggplot(mtcars, aes("a", model, fill=mpg)) + 
  geom_tile() + 
    scale_fill_gradient2(midpoint=median(mtcars$mpg), mid="#ffffbf", low="#91bfdb", high="#fc8d59") + 
  # Terrible for color blind people
  #    scale_fill_gradient2(midpoint=median(mtcars$mpg), mid=muted("green")) + 
  ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_hue.png", width=1280/300*(3/4), theme=theme_preso+theme_extras)

ggplot(mtcars, aes("a", reorder(model, mpg), fill=mpg)) + 
  geom_tile() + 
    scale_fill_gradient2(midpoint=median(mtcars$mpg), mid="#ffffbf", low="#91bfdb", high="#fc8d59") + 
  # Terrible for color blind people
  #    scale_fill_gradient2(midpoint=median(mtcars$mpg), mid=muted("green")) + 
  ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_hue_ordered.png", width=1280/300*(3/4), theme=theme_preso+theme_extras)
```

## Saturation

Unordered and ordered plots with saturation.

```{r}
ggplot(mtcars, aes("a", model, fill=mpg)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high=muted("blue"))+ 
  ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_saturation.png", width=1280/300*(3/4), theme=theme_preso+theme_extras)

ggplot(mtcars, aes("a", reorder(model, mpg), fill=mpg)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high=muted("blue"))+ 
  ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_saturation_ordered.png", width=1280/300*(3/4), theme=theme_preso+theme_extras)
```

For accurate ratioing with saturation, the scale has to extend to zero.

```{r}
ggplot(mtcars, aes("a", reorder(model, mpg), fill=mpg)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high=muted("blue"))+ 
  ylab(NULL) + theme_extras + expand_limits(fill=0)
preso_save("~/hhsd_images/mtcars_saturation_expanded.png", width=1280/300*(3/4), theme=theme_preso+theme_extras)
```


## Area

```{r}
mtcars <- mtcars %>%
  mutate(model=reorder(model, mpg)) %>%
  arrange(mpg, model)

theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank())

cols<-4
ggplot(mtcars, aes(rep(1:cols, 32/cols), rep(1:(32/cols), each=cols))) + 
  geom_point(aes(size=mpg), color="white") + 
  geom_text(aes(label=model), color="white", position=position_nudge(y=0.5), size=2) +
  expand_limits(x=c(0.5,4.5), y=c(1,8.5))+ ylab(NULL) + scale_size_area() + theme_extras 
preso_save("~/hhsd_images/mtcars_area.png", theme=theme_preso+theme_extras)
```

Again, accurate ratioing depends on a scale that goes to zero.

```{r}
ggplot(mtcars, aes(rep(1:cols, 32/cols), rep(1:(32/cols), each=cols))) + 
  geom_point(aes(size=mpg), color="white") + 
  geom_text(aes(label=model), color="white", position=position_nudge(y=0.5), size=2) +
  expand_limits(x=c(0.5,4.5), y=c(1,8.5), size=15)+ ylab(NULL) + theme_extras 
preso_save("~/hhsd_images/mtcars_area_nonzero.png", theme=theme_preso+theme_extras)
```

"Bubble charts" use size to indicate the weight of a point.  In these plots the outliers have low weight (small size).

```{r}
to_plot<-data.frame(p=runif(100),
                    n=rpois(100, 5)^2) %>%
  mutate(y=rbinom(100,n,p))
ggplot(to_plot,aes(p,y/n,size=n)) +geom_point(color="white") +scale_size_area()
ggplot(to_plot,aes(p,y/n)) +geom_point(color="white")

to_plot<-data.frame(trial=1:100,
                    n=rpois(100, 5)^2) %>% 
  mutate(y=rbinom(100,n,0.5)/n)
ggplot(to_plot, aes(trial, y)) + geom_point(color="white")
preso_save("~/hhsd_images/flips_same.png")
ggplot(to_plot, aes(trial, y, size=n)) + geom_point(color="white", alpha=2/3) + scale_size_area()
preso_save("~/hhsd_images/flips_area.png")
```

## Angle

Again, note that the scale goes to zero (a horizontal line).

```{r}
make_segment <- function(df, scale_low, scale_high) {
  scale_range <- scale_high-scale_low
  fraction <- (df$mpg-scale_low) / scale_range
  theta <- fraction * pi/2
  return(data.frame(x=c(0, cos(theta)),
                    y=c(0, sin(theta))))
}
mtcars_angle <- mtcars %>%
  group_by(model) %>%
  do(make_segment(., 0, max(mtcars$mpg)))
#  do(make_segment(., min(mtcars$mpg), max(mtcars$mpg)))

theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank(),
                      strip.text.x = element_text(size=rel(0.5), hjust=0))

ggplot(mtcars_angle, aes(x,y,group=model))+
  geom_path(color="white")+
  facet_wrap(~model,ncol=8)+coord_equal() +
  theme_extras
preso_save("~/hhsd_images/mtcars_slope.png", theme=theme_preso+theme_extras)
```

## Length

In the presentation, I use the jittered bars, so that the audience can't rely on a common scale (which is next in the hierarchy).

```{r}
theme_extras <- theme(axis.text.y=element_text(size=rel(0.7)),
                      axis.title.x=element_blank(),
                      axis.text.x=element_blank(),
                      axis.ticks.x=element_blank())

mtcars$random <- runif(32,0,5)
ggplot(mtcars, aes(x=random, xend=mpg+random, y=model, yend=model)) +
  geom_segment(size=2, color="white") + xlab(NULL) + ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_length.png", theme=theme_preso+theme_extras)


theme_extras <- theme(axis.text.y=element_text(size=rel(0.7)))

ggplot(mtcars, aes(x=0, xend=mpg, y=model, yend=model)) +
  geom_segment(size=2, color="white") + xlab(NULL) + ylab(NULL) + theme_extras
preso_save("~/hhsd_images/mtcars_bars.png", theme=theme_preso+theme_extras)
```

Error bars encode information using length not on a common scale.

```{r}
trials<-10
flips<-data.frame(trial=1:trials) %>%
  mutate(n=round(runif(trials, 20, 500)),
         y=rbinom(trials, n, 0.5),
         p=y/n,
         se=sqrt(p*(1-p)/n),
         lower=p-1.96*se,
         upper=p+1.96*se)
ggplot(flips, aes(trial, p, ymin=lower, ymax=upper)) + geom_pointrange(color="white")
preso_save("~/hhsd_images/flips.png")
```

## Position on non-aligned scales

Doing this with facets is the easiest way in ggplot, but uses too much space.

```{r}
theme_extras <- theme(axis.title.y=element_blank(),
                      axis.text.y=element_blank(),
                      axis.ticks.y=element_blank(),
                      strip.text.x = element_text(size=rel(0.5), hjust=0),
                      panel.grid.major =   element_line(colour = "grey50", size = 0.2),
                      panel.grid.minor =   element_line(colour = "grey50", size = 0.2, linetype="dotted")
                      )

ggplot(mtcars, aes(x=mpg, y="a")) + geom_point(color="white", size=1) + 
  scale_x_continuous(breaks=c(10,20,30)) +
  facet_wrap(~model, ncol=8) + theme_extras
preso_save("~/hhsd_images/mtcars_nonaligned.png", theme=theme_preso+theme_extras)
```

## Position on a common scale

Ahhhh...

```{r}
theme_extras <- theme(axis.text.y=element_text(size=rel(0.7)),
                      axis.title.y=element_text(size=rel(0.7)),
                      panel.grid.major =   element_line(colour = "grey20", size = 0.2))

ggplot(mtcars, aes(x=mpg, y=model)) + 
  geom_point(color="white") + 
  theme_extras + ylab(NULL)
preso_save("~/hhsd_images/mtcars_aligned.png", theme=theme_preso+theme_extras)
```

Ordering still crucial.

```{r}
mtcars$model2<-factor(as.character(mtcars$model))
ggplot(mtcars, aes(x=mpg, y=model2)) + 
  geom_point(color="white") + 
  theme_extras + ylab(NULL)
preso_save("~/hhsd_images/mtcars_aligned_alpha.png", theme=theme_preso+theme_extras)
```

Dropping the scale to zero is a mistake here, lowers resolution and isn't needed for accurate ratioing.

```{r}
ggplot(mtcars, aes(x=mpg, y=model)) + 
  geom_point(color="white") + 
  theme_extras + ylab(NULL) + expand_limits(x=0)
preso_save("~/hhsd_images/mtcars_aligned_zero.png", theme=theme_preso+theme_extras)
```


## Observations

### Stacked anything is nearly always a mistake


A stacked bar chart does two things, neither of the well.

```{r}
ggplot(diamonds, aes(clarity, fill=cut, group=cut)) + 
  geom_bar(stat="count", position="stack") +
  scale_y_continuous("Count", labels=comma)
preso_save("~/hhsd_images/diamonds_stack.png")
```

The first is the total in each group.

```{r}
diamonds_summary<-diamonds %>%
  group_by(clarity) %>%
  summarize(n=length(clarity))

ggplot(diamonds_summary, aes(clarity, n)) + 
  geom_line(group=1, color="white") + 
  geom_point(color="white") + 
  expand_limits(y=0) +
  scale_y_continuous("Count", labels=comma)
preso_save("~/hhsd_images/diamonds_overall.png")
```

The second in the total in each combination, which is better shown with a parallel coordinates plot.

```{r}
diamonds_summary2<-diamonds %>%
  group_by(clarity, cut) %>%
  summarize(n=length(clarity))

ggplot(diamonds_summary2, aes(clarity, n, color=cut, group=cut)) + 
  geom_line() + 
  geom_point() + 
  expand_limits(y=0) +
  scale_y_continuous("Count", labels=comma)
preso_save("~/hhsd_images/diamonds_freqpoly.png")
```

Some phone data (from [wikipedia](https://en.wikipedia.org/wiki/Mobile_operating_system)).

```{r}
phone_data<-
"Quarter	Windows Mobile	RIM	Symbian	iOS	Android	Bada	Windows Phone	Other
2016 Q2[39]	0	400.4	0	44395	296912	0	1971	680.6
2016 Q1[40]	0	660	0	51630	293771	0	2400	791
2015 Q4[41]	0	906.9	0	71526	325394	0	4395	887.3
2015 Q3[42]	0	977	0	46062	298797	0	5874	1133.6
2015 Q2[43]	0	1153	0	48086	271010	0	8198	1229
2015 Q1[44]	0	1325	0	60177	265012	0	8271	1268
2014 Q4[45]	0	1734	0	74832	279058	0	10425	1286.9
2014 Q3[46]	0	2420	0	38187	254354	0	9033	1310.2
2014 Q2[47]	0	2044	0	35345	243484	0	8095	2044
2014 Q1[44]	0	1714	0	43062	227549	0	7580	1371
2013 Q4[48]	0	1807	0	50224	219613	0	8534	1994
2013 Q3[49]	0	4401	458	30330	205023	633	8912	475
2013 Q2[50]	0	6180	631	31900	177898	838	7408	472
2013 Q1[51]	0	6219	1349	38332	156186	1371	5989	600
2012 Q4[52]	0	7333	2569	43457	144720	2684	6186	713
2012 Q3[53]	0	8947	4405	23550	122480	5055	4058	684
2012 Q2[54]	0	7991	9072	28935	98529	4209	4087	863
2012 Q1[55]	0	9939	12467	33121	81067	3842	2713	1243
2011 Q4[56]	0	13185	17458	35456	75906	3111	2759	1167
2011 Q3[57]	0	12701	19500	17295	60490	2479	1702	1018
2011 Q2[58]	0	12652	23853	19629	46776	2056	1724	1051
2011 Q1	982	13004	27599	16883	36350	1862	1600	1495
2010 Q4[56]	3419	14762	32642	16011	30801	2027	0	1488
2010 Q3[57]	2204	12508	29480	13484	20544	921	0	1991
2010 Q2[58]	3059	11629	25387	8743	10653	577	0	2011
2010 Q1[59]	3696	10753	24068	8360	5227	0	0	2403
2009 Q4[60]	4203	10508	23857	8676	4043	0	0	2517
2009 Q3[61]	3260	8523	18315	7040	1425	0	0	2531
2009 Q2[62]	3830	7782	20881	5325	756	0	0	2398
2009 Q1[63]	3739	7534	17825	3848	575	0	0	2986
2008 Q4[64]	4714	7443	17949	4079	639	0	0	3319
2008 Q3[65]	4053	5800	18179	4720	0	0	0	3763
2008 Q2[66]	3874	5594	18405	893	0	0	0	3456
2008 Q1[64]	3858	4312	18400	1726	0	0	0	4113
2007 Q4[64]	4374	4025	22903	1928	0	0	0	3536
2007 Q3[65]	4180	3192	20664	1104	0	0	0	3612
2007 Q2[66]	3212	2471	18273	270	0	0	0	3628
2007 Q1[64]	2931	2080	15844	0	0	0	0	4087"
phones<-read.delim(textConnection(phone_data))
phones <- phones %>%
  mutate(Other = Other + Bada + Windows.Phone,
         Bada = NULL,
         Windows.Phone = NULL,
         Quarter = str_replace(Quarter, "\\[\\d+\\]", ""),
         Year = as.integer(str_split_fixed(Quarter, " ", 2)[,1]),
         Quarter = str_split_fixed(Quarter, " ", 2)[,2],
         qtr = as.integer(str_replace(Quarter, "Q", "")),
         year = Year+0.25*(qtr-1)) %>%
  gather(os, ct, -Year, -Quarter, -qtr, -year) %>%
  mutate(ct = as.numeric(ct)) %>%
  group_by(year) %>%
  mutate(share = ct/sum(ct))
phones
```

All rules have exceptions.  A stacked area chart with phone data is ok because it makes the point more effectively.

```{r}
ggplot(phones, aes(year, share, group=os, fill=os)) + geom_area()
preso_save("~/hhsd_images/phones_area.png")
```

The line chart is weaker.

```{r}
ggplot(phones, aes(year, share, group=os, color=os)) + geom_line()
```

### Growth charts usually aren't

This stacked area chart sucks.

```{r}
library(gapminder)
continent_pop <- gapminder %>%
  group_by(continent, year) %>%
  summarize(pop=sum(as.numeric(pop))) %>%
  ungroup()
ggplot(continent_pop, aes(year, pop/1e9, group=continent, fill=continent)) + 
  geom_area() + 
  scale_y_continuous("Population (Billions)", labels=comma) +
  scale_color_brewer(type="qual", palette=1) + xlab(NULL)
preso_save("~/hhsd_images/population_area.png")
```

This line chart isn't much better if you care about growth rates, becuse growth is encoded with angle (slope).

```{r}
ggplot(continent_pop, aes(year, pop/1e9, group=continent, color=continent)) + 
  geom_line() + scale_y_log10("Population (Billions)", labels=comma) +
  scale_color_brewer(type="qual", palette=1) + xlab(NULL)
preso_save("~/hhsd_images/population.png")
```

If growth is important, take derivatives and plot growth directly on the y-axis.

```{r}
continent_pop <- continent_pop %>%
  group_by(continent) %>%
  arrange(year) %>%
  mutate(growth=c(NA,diff(pop)))
ggplot(continent_pop, aes(year, growth/pop, group=continent, color=continent)) + 
  geom_line() + scale_color_brewer(type="qual", palette=1) + 
  xlab(NULL) + scale_y_continuous("Pct. Growth", labels=percent)
preso_save("~/hhsd_images/population_growth.png")
```

### Putting the data on a common scale helps tremendously

```{r}
host_dat = data.frame(t=seq.POSIXt(as.POSIXct("2016-06-01 00:00:00"),
                                   as.POSIXct("2016-06-01 01:00:00"),
                                   by=30)) %>%
  mutate(cpu=rnorm(length(t), 40, 5),
         latency=rnorm(length(t), 4, 0.1))

host_dat$cpu[50] = 100
host_dat$latency[60] = host_dat$latency[60]*3

host_dat_w <- host_dat %>%
  gather(var, value, -t) %>%
  group_by(var) %>%
  mutate(std_value = (value-mean(value))/sd(value))
```

The user has to work to see if the peaks are aligned.

```{r}
ggplot(host_dat_w, aes(t, value)) + geom_line(color="white") + 
  facet_wrap(~var, scales="free") + xlab(NULL)
preso_save("~/hhsd_images/cpu_latency_nonaligned.png")
```

You want to put them on the same scale, but the magnitude of the series is different.

```{r}
ggplot(host_dat_w, aes(t, value, color=var, group=var)) + 
  geom_line() + xlab(NULL) + ylab("Value") 
preso_save("~/hhsd_images/cpu_latency_aligned.png")
```

So "standardize" them by suntracting the mean and dividing by the standard deviation.

```{r}
ggplot(host_dat_w, aes(t, std_value, color=var, group=var)) + 
  geom_line() + xlab(NULL) + ylab("Standardized Value")
preso_save("~/hhsd_images/cpu_latency_std_aligned.png")
```

TODO: Show deseasonalizing too (usually easy and automatic if many seasonal cycles are available).

### Scatter plots show relationships directly

```{r}
dat <- data.frame(x=seq(1,6*pi,length.out = 100))
# Two series, noe of which varies with the square, the other linearly
dat <- dat %>% 
  mutate(Dependency=1.4+sin(x), 
         Service=(1.1+sin(x))^2)
# Jitter the points
dat <- dat %>%
  mutate(Service=rnorm(length(x), mean = Service, sd=Service/10),
         Dependency=rnorm(length(x), mean = Dependency, sd=Dependency^(2/3)/10))
# Put in long format
dat_l <- dat %>%
  gather(var, value, -x)
```

They move together, but what, preciscely is the nature of the relationship?

```{r}
ggplot(dat_l, aes(x, value))+geom_line(color="white")+facet_wrap(~var, scales="free")
preso_save("~/hhsd_images/ts_vs_scatter_facet.png")
```

```{r}
ggplot(dat_l, aes(x, value, color=var))+geom_line()
preso_save("~/hhsd_images/ts_vs_scatter_ts.png")
```

A scatterplot is more revealing.

```{r}
ggplot(dat, aes(Dependency, Service)) + geom_point(color="white") + 
  ylab("Some service's latency") + xlab("Its dependency")
preso_save("~/hhsd_images/ts_vs_scatter_scatter.png")
```

It's clear that the relationship isn't linear.

```{r}
ggplot(dat, aes(Dependency, Service))+geom_point(color="white")+geom_smooth(method="lm") + 
  ylab("Some service's latency") + xlab("Its dependency")
preso_save("~/hhsd_images/ts_vs_scatter_lm.png")
```

A loess fit looks nice.

```{r}
ggplot(dat, aes(Dependency, Service))+geom_point(color="white")+geom_smooth(method="loess") + 
  ylab("Some service's latency") + xlab("Its dependency")
preso_save("~/hhsd_images/ts_vs_scatter_loess.png")
```


# Part 3: Assembly

## The Law of Continuity

The circle from the start is a good example, as is a simple scatterplot.  Note that the continuity just has to be suggestive, not perfect.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color="white") + 
  xlab("weight (1,000 lbs)")
preso_save("~/hhsd_images/wt_mpg_scatter.png")
```

### The Law of Similarity

Color is pretty good.

```{r}
# Make number of cylinders a factor, so ggplot will choose a discrete scale.
mtcars$cyl<-factor(mtcars$cyl)

ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(color=cyl)) + 
  xlab("weight (1,000 lbs)") +
  scale_color_brewer("Cylinders", type="qual", palette=1)
preso_save("~/hhsd_images/wt_mpg_scatter_color.png")
```

These filled shapes aren't great.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), color="white") + 
  xlab("weight (1,000 lbs)") +
  scale_shape("Cylinders")
preso_save("~/hhsd_images/wt_mpg_scatter_shape1.png")
```

But these open ones are great.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), color="white") + 
  xlab("weight (1,000 lbs)") +
  scale_shape_manual("Cylinders", values = c(1,2,3))
preso_save("~/hhsd_images/wt_mpg_scatter_shape2.png")
```

It's perfectly ok to redundantly encode using shape and color.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl, color=cyl)) + 
  xlab("weight (1,000 lbs)") +
  scale_shape_manual("Cylinders", values = c(1,2,3)) +
  scale_color_brewer("Cylinders", type="qual", palette=1)
preso_save("~/hhsd_images/wt_mpg_scatter_both.png")
```

You'd think this would be a great encoding, but it's terrible because of the similar curvature of the sixes and eights.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl), color="white", size=2) + 
  xlab("weight (1,000 lbs)") +
  scale_shape_manual(values = c(52,54,56))
preso_save("~/hhsd_images/wt_mpg_scatter_letters.png")
```

Overlaying linear fits encloses space and makes the grouped objects very strong.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(aes(shape=cyl, color=cyl)) + 
  geom_smooth(aes(group=cyl, color=cyl), method="lm") + 
  xlab("weight (1,000 lbs)") +
  scale_shape_manual("Cylinders", values = c(1,2,3)) +
  scale_color_brewer("Cylinders", type="qual", palette=1)
preso_save("~/hhsd_images/wt_mpg_scatter_lm.png")
```

### Law of Proximity

A dodged bar chart.  Try to compare the blue vs. green bars.  I find it very difficult.

```{r}
ggplot(diamonds, aes(clarity, fill=cut, group=cut)) + 
  geom_bar(stat="count", position="dodge") +
  scale_y_continuous("Count", labels=comma)
preso_save("~/hhsd_images/diamonds_dodge.png")
```

### Assembly with iris data

Another dataset that's good for demonstrating assembly.  I didnt use these in the presentation.

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species))+geom_jitter()
ggplot(iris, aes(Sepal.Length, Petal.Width, color=Species))+geom_jitter()
ggplot(iris, aes(Sepal.Length, Petal.Length, color=Species))+geom_jitter()
```


# Part 4: Detection

People frequenty hinder detection by lowering the contrast in luminance.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color="gray30") + 
  xlab("weight (1,000 lbs)")
preso_save("~/hhsd_images/wt_mpg_scatter_grey.png")
```

Or by hiding the data with non-data clutter, like gridlines.

```{r}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color="white", size=0.2) + 
  xlab("weight (1,000 lbs)")
preso_save("~/hhsd_images/wt_mpg_scatter_small.png")

theme_extras <- theme(panel.grid.major =   element_line(colour = "grey20", size = 0.2))
p<-ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color="white", size=1.5) + 
  xlab("weight (1,000 lbs)")

theme_extras <- theme(panel.border =       element_rect(fill = NA, colour = "white"),
                      panel.grid.major =   element_line(colour = "white", size = 0.2))
p+theme_extras
preso_save("~/hhsd_images/wt_mpg_scatter_grid1.png", theme=theme_preso + theme_extras)

theme_extras <- theme(panel.border =       element_rect(fill = NA, colour = "white"),
                      panel.grid.minor =   element_line(colour = "white", size = 0.2),
                      panel.grid.major =   element_line(colour = "white", size = 0.5))
p+theme_extras
preso_save("~/hhsd_images/wt_mpg_scatter_grid2.png", theme=theme_preso + theme_extras)

theme_extras <- theme(panel.border =       element_rect(fill = NA, colour = "white"),
                      panel.grid.minor =   element_line(colour = "white", size = 0.5),
                      panel.grid.major =   element_line(colour = "white", size = 1))
p+theme_extras
preso_save("~/hhsd_images/wt_mpg_scatter_grid4.png", theme=theme_preso + theme_extras)
```

The Hermann Grid illusion.  Do you see the faint grey dots?

```{r}
theme_extras <- theme(panel.border =       element_rect(fill = NA, colour = "white"),
                      panel.grid.major =   element_line(colour = "white", size = 3.5))
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color="white", size=0, alpha=0) + 
  xlab("weight (1,000 lbs)") + theme_extras
preso_save("~/hhsd_images/wt_mpg_scatter_grid3.png", theme=theme_preso + theme_extras)
```

Decent grid and borders

```{r}
theme_extras <- theme(panel.border =       element_rect(fill = NA, colour = "grey70"),
                      panel.grid.major =   element_line(colour = "grey70", size = 0.2))
p+theme_extras
preso_save("~/hhsd_images/wt_mpg_scatter_grid5.png", theme=theme_preso + theme_extras)
```

# Part 5: Other Useful results

## Weber's Law

```{r}
weber <- data.frame(bar = c("c", "a", "b", "bb"),
                    zpad = c(3, 0, 0, 0),
                    ylen = c(10, 10.5, 0, 0),
                    xtop = c(2, 1.5, 0, 0))

weber1 <- weber %>% 
  select(-xtop) %>%
  gather(var, val, -bar) 

theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank(),
                      legend.position="none")
```

It's hard to tell if the bars are the same legnth or different.

```{r}
ggplot(weber1, aes(bar, val, fill=var)) + geom_bar(stat="identity") + 
  scale_fill_manual(values=c("white", "black")) +
  expand_limits(y=c(0,15)) + theme_extras
preso_save("~/hhsd_images/weber1.png", theme=theme_preso+theme_extras)
```

But with the little blue ones, it's easy, even though the absolute difference in length is exactly the same as with the white bars (0.5 units).

```{r}
weber2 <- weber %>% 
  gather(var, val, -bar) 

ggplot(weber2, aes(bar, val, fill=var)) + geom_bar(stat="identity") + 
  scale_fill_manual(values=c(muted("blue",l=60), "white", "black")) +
  expand_limits(y=c(0,15)) + theme_extras
preso_save("~/hhsd_images/weber2.png", theme=theme_preso+theme_extras)
```

Define eight different curves.

```{r}
to_plot<-data.frame(x=seq(1,100,length.out=500)) %>%
  mutate(y=log(x),
         y2 = -(((x-100))^2),
         y03 = 100-1/x*y,
         y04 = 100-2/x*y,
         y05 = 100-2/(x+10)*y,
         y06 = 100-2/(x+5)*y,
         y07 = 99.8-1.5/(x+5)*y,
         y08 = 100-2.5/(x+5)*y,
         y09 = 100-1.5/(x+52)*y,
         y10 = 100-1.5/(x+25)*y
         )
```

Without gridlines it's hard to compare them precisely.

```{r}
theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank(),
                      panel.border = element_rect(fill = NA, colour = "grey70"),
                      strip.text=element_blank())

ggplot(to_plot %>% select(-y2, -y) %>% gather(var, val, -x),
       aes(x,val)) + geom_line(color="white") + facet_wrap(~var, ncol=2) + theme_extras
preso_save("~/hhsd_images/weber_grid1.png", width=1280/300/2, theme=theme_extras)
```

It's much easier with gridlines, because of Weber's law.

```{r}
theme_extras <- theme(axis.title=element_blank(),
                      axis.text=element_blank(),
                      axis.ticks=element_blank(),
                      panel.border = element_rect(fill = NA, colour = "grey70"),
                      panel.grid.major = element_line(colour = "grey60", size = 0.2),
                      strip.text=element_blank())

ggplot(to_plot %>% select(-y2, -y) %>% gather(var, val, -x),
       aes(x,val)) + geom_line(color="white") + facet_wrap(~var, ncol=2) + theme_extras
preso_save("~/hhsd_images/weber_grid2.png", width=1280/300/2, theme=theme_extras)
```

## You are most sensitive to changes in angle close to 45 degrees

The same data ploteed three ways.  It's easiest to detect the change in slope near 45 degrees.

```{r}
dat<-data.frame(ct=c(1e6-10000, 1e6, 1e6+15000),
                dt=as.Date(c("2016-06-01","2016-06-02", "2016-06-03")))
p<-ggplot(dat, aes(dt, ct)) + 
  geom_line(group=1, color="white") + 
  geom_point(color="white") + ylab(NULL) + scale_x_date(NULL, labels=date_format("%b-%d"))
p
preso_save("~/hhsd_images/45deg.png")
p + expand_limits(y=c(800000,1200000))
preso_save("~/hhsd_images/45deg_tall.png")
p + expand_limits(x=as.Date(c("2016-05-01","2016-07-01")))
preso_save("~/hhsd_images/45deg_wide.png")
```

Difficult to see with bars (where the y-axis must extend to zero).

```{r}
ggplot(dat, aes(dt, ct)) + 
  geom_bar(stat="identity", fill="white") + 
  ylab(NULL) + scale_x_date(NULL, labels=date_format("%b-%d"))
preso_save("~/hhsd_images/45deg_bars.png", width=1280/300/2)
```

For banking to 45, see the example with the sunspots data in part 6.

## The eye measures perdendicular distance

Where is the distance between the lines the greatest?

```{r}
dat<-data.frame(x=seq(-10, 8, length.out = 100)) %>%
  mutate(y=-2*(x^2)+300,
         y2=y-seq(30,5,length.out = 100)) 

dat_w <- dat %>%
  gather(var, val, y, y2)

ggplot(dat_w, aes(x, val, group=var))+
  geom_line(color="white")
preso_save("~/hhsd_images/gap_lines.png", width=1280/300*(2/3))
```

Here's the distance plotted directly.  Suprisingly, the distance is the largest at the left edge.

```{r}
ggplot(dat, aes(x, y-y2)) + geom_line(color="white") + ylab("gap")
preso_save("~/hhsd_images/gap_direct.png")
```


## Y-axis to zero?

See the saturation, area and length examples from the estimation section with mtcars.

With bars the y-axis must go to zero.  This obscures the variation in this dataset.

```{r}
to_plot<-data.frame(x=1:20) %>%
  mutate(y=1e6 + 100*x*(-1)^x)
ggplot(to_plot, aes(x,y)) + geom_bar(fill="white",stat="identity") + xlab(NULL) + ylab(NULL)
preso_save("~/hhsd_images/yaxis_bars.png")
```

If we switch to points (and position on a common scale), we are free to clip the y-axis to just the relevant interval.

```{r}
ggplot(to_plot, aes(x,y)) + geom_line(color="white") + geom_point(color="white") + xlab(NULL) + ylab(NULL)
preso_save("~/hhsd_images/yaxis_lines.png")
```


## Hue (used carefully) is good for revealing fine structure

```{r}
ramps<-data.frame(x=seq(-5,5,length.out=200)) %>%
  mutate(linear=x/10+0.5,
         sigmoid=1/(1+exp(-x)))

theme_extras <- theme(axis.title.y=element_blank(),
                      axis.text.y=element_blank(),
                      axis.ticks.y=element_blank(),
                      legend.position="none")

ggplot(ramps, aes(x,1,fill=sigmoid)) + geom_tile() + theme_extras
preso_save("~/hhsd_images/ramp_sigmoid.png", theme=theme_preso+theme_extras)

ggplot(ramps, aes(x,1,fill=linear)) + geom_tile() + theme_extras
preso_save("~/hhsd_images/ramp_linear.png", theme=theme_preso+theme_extras)

theme_extras <- theme(axis.title.y=element_blank(),
                      axis.text.y=element_blank(),
                      axis.ticks.y=element_blank())

ggplot(ramps %>% gather(series,value,-x), aes(x,value,color=series)) + geom_line() + theme_extras
preso_save("~/hhsd_images/ramp_both_lines.png", theme=theme_preso+theme_extras)

theme_extras <- theme(axis.title.y=element_blank(),
                      legend.position="none")

ggplot(ramps %>% gather(series,value,-x), aes(x,series,fill=value)) + geom_tile() + theme_extras
preso_save("~/hhsd_images/ramp_both.png", theme=theme_preso+theme_extras)

jet.colors <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan", "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
ggplot(ramps %>% gather(series,value,-x), aes(x,series,fill=value)) + geom_tile() + theme_extras + scale_fill_gradientn(colors = jet.colors(7))
preso_save("~/hhsd_images/ramp_both_hue.png", theme=theme_preso+theme_extras)
```

Viridis not a good as colorjet.  Presumably because colorjet traverses more of the space of colors.

```{r}
ggplot(ramps %>% gather(series,value,-x), aes(x,series,fill=value)) + geom_tile() + theme_extras + scale_fill_viridis()
```


# Part 6: Tools/Workflow

## Sunspot data

The data is from: [http://www.sidc.be/silso/DATA/SN_d_tot_V2.0.txt](http://www.sidc.be/silso/DATA/SN_d_tot_V2.0.txt).

First we load it in and tidy it up.

```{r}

# Format:
#Line format [character position]:
#- [1-4] Year
#- [6-7] Month
#- [9-10] Day
#- [12-19] Decimal date
#- [22-24] Daily sunspot number
#- [26-30] Standard deviation
#- [33-35] Number of observations
#- [37] Definitive/provisional indicator

positions <- fwf_positions(start=c(1,6,9,12,22,26,33,37), 
                           end=c(4,7,10,19,24,30,35,37),
                           col_names=c("year", "month", "day", "decimal_date", 
                                       "sunspot_number", "std_dev", "n", "provisional"))
sunspots_raw<-read_fwf("~/Dropbox/Velocity Amsterdam 2016/SN_d_tot_V2.0.txt", col_positions=positions)
sunspots <- sunspots_raw %>%
  mutate(day = as.integer(day),
         dt = as.Date(sprintf("%4d-%02d-%02d", year, month, day)),
         sunspot_number = ifelse(sunspot_number==-1, NA, as.numeric(sunspot_number)),
         std_dev = ifelse(std_dev==-1, NA, std_dev),
         provisional = !is.na(provisional)) %>%
  # There are many missing values before 1850 which complicate analysis
  filter(dt >= "1850-01-01") %>%
  select(-year, -month, -day, -decimal_date)
summary(sunspots)
```

A first attempt at plotting... yuck.

```{r}
ggplot(sunspots, aes(dt, sunspot_number)) + 
  geom_line(color="white") + 
  xlab(NULL) + ylab("Sunspot Number")
preso_save("~/hhsd_images/sunspots_line.png")
```

Plot as dots, use alpha transparency to reduce overplotting.  Notice the gaps between zero and non-zero.  You wouldn't have noticed those without visualization.

```{r}
ggplot(sunspots, aes(dt, sunspot_number)) + 
  geom_point(size=0.5, alpha=1/40, color="white") + 
  xlab(NULL) + ylab("Sunspot Number")
preso_save("~/hhsd_images/sunspots_point.png")
```

There are lots of things we could try to explain about this data, but the most obivous one is the major cycle that's about 11 years long.

```{r}
minima <- as.Date("1954-01-01") + seq(-9, 5) * 11 * 365.25
ggplot(sunspots, aes(dt, sunspot_number)) + 
  geom_point(size=0.5, alpha=1/40, color="white") + 
  geom_vline(xintercept=as.numeric(minima), color="lightblue") +
  xlab(NULL) + ylab("Sunspot Number")
preso_save("~/hhsd_images/sunspots_vlines.png")
```

To make it easier to work with, let's take monthly averages of the data.

```{r}
# This works, but it's silly to work on this data at a daily level
#sunspots_ts<- ts(sunspots$sunspot_number, frequency=11*365.25)
#z <- fourier(sunspots_ts, K=3)
#fit <- auto.arima(sunspots_ts, d=1, xreg=z, allowdrift = F, seasonal=FALSE, trace = T)

sunspots_month <- sunspots %>% 
  mutate(month = as.Date(format(dt, "%Y-%m-01"))) %>%
  group_by(month) %>%
  summarize(mean_sn = mean(sunspot_number, na.rm=T))
summary(sunspots_month)

ggplot(sunspots_month, aes(month, mean_sn)) + geom_line(color="white")
preso_save("~/hhsd_images/sunspots_month.png")
#ggplot(sunspots_month, aes(month, mean_sn)) + geom_point(color="white")
```

Now we can put a loess smoother over the monthly data, and we finally have a nice visualization of the series.

```{r}
ggplot(sunspots_month, aes(month, mean_sn)) + geom_point(color="white", alpha=1/3) + 
  geom_smooth(method="loess", span=0.05, n=400) + 
  xlab(NULL) + ylab("Mean Sunspot Number")
preso_save("~/hhsd_images/sunspots_month_smooth.png")
```

Now let's impose a model on the data.

```{r, cache=TRUE, autodep=TRUE}
library(forecast)
tb<-tbats(sunspots_month$mean_sn, 
          use.trend=F, use.box.cox=F, 
          seasonal.periods = c(12*11))
tb_components_mat<-tbats.components(tb)
tb_components<-data.frame(month=sunspots_month$month,
                          observed = tb_components_mat[,"observed"],
                          level = tb_components_mat[,"level"],
                          season = tb_components_mat[,"season"],
                          residuals = residuals(tb),
                          fitted = fitted(tb))

tb_components_w <- tb_components %>%
  gather(variable, value, -month) %>%
  mutate(variable = factor(variable, levels=c("observed", "level", "season", "fitted", "residuals")))
```

... and visualize the results of the model.

```{r}
ggplot(tb_components_w %>% filter(variable %in% c("observed","level","season")), 
       aes(month, value)) + 
  geom_line(color="white") + 
  facet_wrap(~variable, ncol=1, scales="free_y")
preso_save("~/hhsd_images/sunspots_tbats.png")
```

But is the model good?  A common model diagnostic is to look at the "residuals" the difference between the observed data and the model.  According to our model, this should be normally distributed with a mean of zero.

```{r}
binw=5
ggplot(tb_components, aes(residuals)) + geom_histogram(binwidth=binw, fill="lightgrey")
preso_save("~/hhsd_images/sunspots_residuals.png")
```

If you don't have great intuition about what a normal curve looks like, it can help to overlay one.  It look like our model ok, but not great.

```{r}
curve = data.frame(x=seq(min(tb_components$residuals), max(tb_components$residuals), length.out = 200)) %>%
  mutate(y=(binw*length(tb_components$residuals)) * dnorm(x, mean(tb_components$residuals), sd(tb_components$residuals)))
ggplot(tb_components, aes(residuals)) + geom_histogram(binwidth=binw, fill="lightgrey") +
  geom_line(data=curve, aes(x,y), color="red")

```


```{r, eval=F}
# Other diagnostic plots
qqnorm(residuals(tb))
plot(acf(residuals(tb), lag.max=150))
```

```{r, eval=F}
# 11 year moving average
plot(ma(sunspots$sunspot_number, round(365.25*11)))
```


Demonstrates banking to 45.  There seems to be a pattern of a sharper attack and slower decay.

```{r}
sunspots_month <- sunspots_month %>%
#  mutate(piece = ifelse(month < '1900-01-01', "1850-1900", ifelse(month < '1950-01-01', "1900-1950", "1950-2000"))) %>% 
  mutate(piece = ifelse(month < '1905-01-01', "1850-1905", ifelse(month < '1960-01-01', "1906-1960", "1960-2015"))) %>% 
  group_by(piece) %>%
  mutate(obs=1:length(month)) %>%
  ungroup()

theme_extras <- theme(axis.title.x=element_blank(),
                      axis.text.x=element_blank(),
                      axis.ticks.x=element_blank())

ggplot(sunspots_month %>% filter(month < '2015-01-01'), 
       aes(obs, mean_sn)) + geom_point(color="white", alpha=1/3) + 
  geom_smooth(method="loess", span=0.05*2, n=150) + facet_wrap(~piece, ncol=1) + 
  theme_extras + 
  ylab("Mean Sunspot Number")

preso_save("~/hhsd_images/sunspots_month_wide.png", theme=theme_preso + theme_extras)
```


# Appendix

## Barley dataset

```{r}
library(lattice) # For the barley dataset
# If you want to encourage comparing between years, then put them on a common scale
ggplot(barley, aes(yield, variety, color=year, shape=year))+
  geom_point()+facet_wrap(~site)
# If you want to encourage comparing between sites, then put them on a common scale
ggplot(barley, aes(variety, yield, group=site, color=site, shape=site))+
  geom_point()+geom_line()+facet_wrap(~year)+coord_flip()

# Can you spot the anomaly?
barley_differences <- barley %>%
  group_by(variety, site) %>%
  arrange(year) %>%
  summarize(difference = yield[2]-yield[1])

ggplot(barley_differences, 
       aes(difference, variety))+
  geom_point(color="white") + facet_wrap(~site)

ggplot(barley_differences, 
       aes(difference, variety))+
  geom_point(color="white") + geom_vline(xintercept=0) + facet_wrap(~site)

ggplot(barley_differences, 
       aes(difference, variety, color=factor(sign(difference))))+
  geom_point()+facet_wrap(~site)
```


```{r, eval=F, include=F}
# A complete theme...
theme_black <- function(base_size = 12, base_family = "Helvetica") {
    theme(
    line =               element_line(colour = "black", size = 0.5, linetype = 1,
                            lineend = "butt"),
    rect =               element_rect(fill = "white", colour = "black", size = 0.5, linetype = 1),
    text =               element_text(family = base_family, face = "plain",
                            colour = "black", size = base_size,
                            hjust = 0.5, vjust = 0.5, angle = 0, lineheight = 0.9, margin = unit(0.5, "lines"), debug=F),
    axis.text =          element_text(size = rel(0.8), colour = "white", margin = unit(0.5, "lines")),
    strip.text =         element_text(size = rel(0.8), colour = "white", margin = unit(0.5, "lines")),

    axis.line =          element_blank(),
    axis.text.x =        element_text(vjust = 1, margin = unit(0.5, "lines")),
    axis.text.y =        element_text(hjust = 1, margin = unit(0.5, "lines")),
    axis.ticks =         element_line(colour = "white", size = 0.2),
    axis.title =         element_text(colour = "white", margin = unit(0.5, "lines")),
    axis.title.x =       element_text(vjust = 1, margin = unit(0.5, "lines")),
    axis.title.y =       element_text(angle = 90, margin = unit(0.5, "lines")),
    axis.ticks.length =  unit(0.3, "lines"),
    #axis.ticks.margin =  unit(0.5, "lines"),

    legend.background =  element_rect(colour = NA),
    legend.margin =      unit(0.2, "cm"),
    legend.key =         element_rect(fill = "black", colour = "white"),
    legend.key.size =    unit(1.2, "lines"),
    legend.key.height =  NULL,
    legend.key.width =   NULL,
    legend.text =        element_text(size = rel(0.8), colour = "white", margin = unit(0.5, "lines")),
    legend.text.align =  NULL,
    legend.title =       element_text(size = rel(0.8), face = "bold", hjust = 0, colour = "white", margin = unit(0.5, "lines")),
    legend.title.align = NULL,
    legend.position =    "right",
    legend.direction =   "vertical",
    legend.justification = "center",
    legend.box =         NULL,

    panel.background =   element_rect(fill = "black", colour = NA),
    panel.border =       element_rect(fill = NA, colour = "white"),
    panel.grid.major =   element_line(colour = "grey20", size = 0.2),
    panel.grid.minor =   element_line(colour = "grey5", size = 0.5),
    panel.margin =       unit(0.25, "lines"),

    strip.background =   element_rect(fill = "grey30", colour = "grey10"),
    strip.text.x =       element_text(),
    strip.text.y =       element_text(angle = -90),

    plot.background =    element_rect(colour = "black", fill = "black"),
    plot.title =         element_text(size = rel(1.2)),
    plot.margin =        unit(c(1, 1, 0.5, 0.5), "lines"),

    complete = TRUE
  )
}
```


